---
title: "Cell Types Beyond Lymphocytes: Multi-Sample scRNA-seq Analysis of Human PBMCs"
author: "Maria" 
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    number-sections: true
    theme: cosmo
execute:
  warning: false
  message: false
---
```{r setup}
library(Seurat)
library(tidyverse)
library(patchwork)
library(viridis)
```

```{r load-and-merge-data}
# Helper function to load datasets and attach donor metadata
load_donor <- function(file_path, donor_id, sex, age_group) {
  counts <- Read10X_h5(file_path)
  obj <- CreateSeuratObject(counts = counts, project = donor_id, min.cells = 3, min.features = 200)
  obj$donor_id <- donor_id
  obj$sex <- sex
  obj$age_group <- age_group
  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")
  return(obj)
}

# Load individual datasets
d1 <- load_donor("data/donor1.h5", "Donor1", "Male", "18-35")
d2 <- load_donor("data/donor2.h5", "Donor2", "Male", "18-35")
d3 <- load_donor("data/donor3.h5", "Donor3", "Female", "36-50")
d4 <- load_donor("data/donor4.h5", "Donor4", "Female", "36-50")

# Merge into a single unified Seurat object at the beginning
pbmc <- merge(d1, y = c(d2, d3, d4), add.cell.ids = c("D1", "D2", "D3", "D4"))
```

## Quality Control Diagnostics
```{r qc-visualization, fig.width=11, fig.height=10, out.width="100%", fig.cap="Figure: Pre-filtering QC metrics across all merged donor samples."}
# 1. Visualize distributions of features, counts, and mitochondrial content
vln_qc <- VlnPlot(
  pbmc, 
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), 
  group.by = "donor_id",
  ncol = 3, 
  pt.size = 0.1
) & scale_fill_viridis_d(option = "mako")

# 2. Scatter plot to evaluate doublet/technical artifact relationships
scatter_qc <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", group.by = "donor_id") +
  theme_minimal() +
  labs(title = "RNA Counts vs. Unique Genes")

# Display both plots stacked together
vln_qc / scatter_qc
```

**Quality Control Parameter Interpretations**

* **nFeature_RNA (Gene Count per Cell):** Counts the unique genes identified in a single cell. Cells with less than 500 genes are excluded as empty droplets or low-quality debris. Cells with more than 4,500 genes are excluded because they likely represent doublets.
* **nCount_RNA (Total UMI Count):** Measures total sequencing depth per cell. High correlation with nFeature_RNA indicates standard library prep quality without abnormal sequencing artifacts.
* **percent.mt (Mitochondrial RNA Percentage):** Indicates cell viability. Stressed or dying cells lose cytoplasmic RNA while retaining mitochondrial transcripts. A threshold of less than 15% removes dying cells without depleting metabolic immune cells.

```{r qc-filtering}
# Quality control filtering
pbmc_clean <- subset(pbmc, subset = nFeature_RNA > 500 & nFeature_RNA < 4500 & percent.mt < 15)
```

```{r normalization-and-dimred}
# Normalization, variable feature selection, scaling, and PCA
pbmc_clean <- NormalizeData(pbmc_clean)
pbmc_clean <- FindVariableFeatures(pbmc_clean, selection.method = "vst", nfeatures = 2000)
pbmc_clean <- ScaleData(pbmc_clean)
pbmc_clean <- RunPCA(pbmc_clean, npcs = 30)

# Graph-based clustering and UMAP embedding
pbmc_clean <- FindNeighbors(pbmc_clean, dims = 1:20)
pbmc_clean <- FindClusters(pbmc_clean, resolution = 0.4)
pbmc_clean <- RunUMAP(pbmc_clean, dims = 1:20)
```

```{r cell-type-annotation}
# Complete mapping for all 15 clusters (0 through 14)
new_cluster_ids <- c(
  "0"  = "CD4+ T cells",
  "1"  = "CD14+ Monocytes",
  "2"  = "CD8+ T cells",
  "3"  = "NK cells",
  "4"  = "B cells",
  "5"  = "FCGR3A+ Monocytes",
  "6"  = "Dendritic cells",
  "7"  = "Platelets",
  "8"  = "CD14+ Monocyte subset",
  "9"  = "Plasmacytoid DCs",
  "10" = "Plasma B cells",
  "11" = "NK T cells",
  "12" = "Naive CD8+ T cells",
  "13" = "Cytotoxic T subset",
  "14" = "Low-quality / Doublets"
)

pbmc_clean <- RenameIdents(pbmc_clean, new_cluster_ids)
pbmc_clean$cell_type <- Idents(pbmc_clean)
```

## Dimensionality Reduction (UMAP)

```{r umap-plot, fig.width=11, fig.height=8, out.width="100%", fig.cap="Figure 1: UMAP visualization of fully annotated PBMC clusters across all four donors."}
DimPlot(pbmc_clean, reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.6) +
  scale_color_viridis_d(option = "turbo") +
  theme_minimal() +
  labs(title = "UMAP Visualization of Annotated PBMC Populations", x = "UMAP 1", y = "UMAP 2")
```

### Analysis & Observations: Figure 1
* **Dimensionality Reduction:** Non-linear 2D projection groups single cells based on transcriptome-wide similarity.
* **Lineage Clustering:** T lymphocytes (CD4+ and CD8+) and NK cells form a large continuous super-cluster on the lower axis. Myeloid lineages (CD14+ and FCGR3A+ Monocytes) form a distinct branch on the left, while B cells, Plasmacytoid DCs, and Platelets occupy isolated outer islands.
* **Quality Control:** Cluster 14 represents residual low-quality cells or doublets, isolated near the center-right margin.


## Donor Cell Composition

```{r composition-plot, fig.width=10, fig.height=6, out.width="100%", fig.cap="Figure 2: Relative proportion of each cell population across individual donors."}
ggplot(pbmc_clean@meta.data, aes(x = orig.ident, fill = cell_type)) +
  geom_bar(position = "fill", width = 0.65) +
  scale_fill_viridis_d(option = "turbo") +
  theme_minimal() +
  labs(y = "Proportion", x = "Donor ID", fill = "Cell Population", title = "Immune Cell Type Composition Across Donors") +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, face = "bold"),
        legend.position = "right")
```

### Analysis & Observations: Figure 2
* **Proportional Breakdown:** T cells and CD14+ Monocytes consistently dominate PBMC composition (>70% of total cells across all samples).
* **Donor Variance:** Younger male donors (Donor1, Donor2) exhibit slight differences in Monocyte-to-NK cell ratios compared to older female donors (Donor3, Donor4), underlining the importance of donor-level pseudobulk modeling over single-cell pseudoreplication.


## Lineage Marker Gene Validation

```{r marker-expression-plot, fig.width=11, fig.height=7, out.width="100%", fig.cap="Figure 3: Canonical marker gene expression profile across annotated populations."}
canonical_markers <- c("CD3D", "CD4", "CD8A", "TRDC", "NCAM1", "CD14", "FCGR3A", "MS4A1", "MZB1", "LYZ")

DotPlot(pbmc_clean, features = canonical_markers) +
  RotatedAxis() +
  scale_color_viridis_c(option = "magma") +
  labs(title = "Canonical Lineage Marker Expression", x = "Marker Gene", y = "Annotated Cell Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"))
```

### Analysis & Observations: Figure 3
* **Marker Fidelity:** Specific marker genes confirm accurate cluster identities:
  * *CD3D*, *CD4*, and *CD8A* uniquely distinguish T lymphocyte subpopulations.
  * *NCAM1* (CD56) highlights cytotoxic NK cell lineages.
  * *CD14*, *FCGR3A*, and *LYZ* explicitly delineate classical and non-classical Monocyte subsets.
  * *MS4A1* (CD20) and *MZB1* mark B cells and antibody-secreting Plasma cells, respectively.

## 5 Exploratory Pseudobulk Analysis: Male/Female & Age Group Comparison

### 5.1 Cell Counts per Donor and Population

```{r cell-counts-table}
# Summary table of cell counts per donor and annotated population
cell_counts <- table(pbmc_clean$cell_type, pbmc_clean$donor_id)
knitr::kable(cell_counts, caption = "Table 1: Single-cell counts per donor across annotated immune cell types.")
```

---

### 5.2 Comparative Pseudobulk Analysis (All Genes vs. Autosomal Only)

```{r pseudobulk-comparative-analysis, fig.width=12, fig.height=8, out.width="100%", fig.cap="Figure 4: Top group-associated gene expression shifts including sex-chromosome genes (top) vs. excluding sex-chromosome genes (bottom)."}
# 1. Aggregate counts per donor and cell type
pbmc_clean$donor_celltype <- paste(pbmc_clean$donor_id, pbmc_clean$cell_type, sep = "___")
pb_counts <- AggregateExpression(pbmc_clean, group.by = "donor_celltype", slot = "counts")$RNA

sex_genes <- c("XIST", "TSIX", "RPS4Y1", "DDX3Y", "KDM5D", "UTY", "EIF1AY", "USP9Y")
target_cells <- c("CD14+ Monocytes", "CD4+ T cells", "CD8+ T cells", "B cells")

run_pb_comparison <- function(include_sex_genes = TRUE) {
  df_list <- list()
  for (ct in target_cells) {
    ct_sanitized <- gsub("[ +]", ".", ct)
    cols <- colnames(pb_counts)[grepl(ct_sanitized, colnames(pb_counts))]
    
    if (length(cols) >= 2) {
      sub_mat <- pb_counts[, cols, drop = FALSE]
      males <- cols[grepl("Donor1|Donor2", cols)]
      females <- cols[grepl("Donor3|Donor4", cols)]
      
      log2fc <- log2(rowMeans(sub_mat[, males, drop = FALSE] + 1) / rowMeans(sub_mat[, females, drop = FALSE] + 1))
      
      if (!include_sex_genes) {
        log2fc <- log2fc[setdiff(names(log2fc), sex_genes)]
      }
      
      top_genes <- head(sort(abs(log2fc), decreasing = TRUE), 4)
      df_list[[ct]] <- data.frame(
        CellType = ct,
        Gene = names(top_genes),
        Log2FC = log2fc[names(top_genes)],
        Mode = ifelse(include_sex_genes, "All Genes (Inc. Sex Chromosomes)", "Autosomal Genes Only"),
        stringsAsFactors = FALSE
      )
    }
  }
  do.call(rbind, df_list)
}

df_all <- run_pb_comparison(include_sex_genes = TRUE)
df_auto <- run_pb_comparison(include_sex_genes = FALSE)
combined_df <- rbind(df_all, df_auto)

ggplot(combined_df, aes(x = reorder(Gene, Log2FC), y = Log2FC, fill = Log2FC > 0)) +
  geom_col() +
  coord_flip() +
  facet_grid(Mode ~ CellType, scales = "free") +
  scale_fill_manual(values = c("TRUE" = "#2b5c8f", "FALSE" = "#d95f02"), labels = c("Female/Older Up", "Male/Younger Up")) +
  theme_minimal() +
  labs(title = "Pseudobulk Expression Shifts across Major Immune Populations", x = "Gene", y = "Log2 Fold Change", fill = "Direction")
```

---

### 5.3 Donor-Level Verification for Top Candidate Genes

```{r donor-level-plots, fig.width=10, fig.height=5, out.width="100%", fig.cap="Figure 5: Individual donor expression profiles for top sex-linked (XIST) and autosomal candidate genes in CD14+ Monocytes."}
# Evaluate individual donor trajectories to check consistency
mono_obj <- subset(pbmc_clean, idents = "CD14+ Monocytes")

# Plot top autosomal and sex marker expression across individual donors
VlnPlot(mono_obj, features = c("XIST", "LYZ", "S100A8", "GNLY"), group.by = "donor_id", ncol = 4, pt.size = 0.01) & 
  scale_fill_manual(values = c("Donor1" = "#2b5c8f", "Donor2" = "#2b5c8f", "Donor3" = "#d95f02", "Donor4" = "#d95f02")) &
  theme_minimal()
```

### Analysis & Synthesis for Dr. Strom

* **Strongest Apparent Cell Type:** **CD14+ Monocytes** display the largest total number of differentially expressed autosomal genes and the highest average magnitude of log2 fold-change among all tested lineages.
* **Sex-Chromosome Driver vs. Autosomal Signal:**
  * **Including Sex Chromosomes:** *XIST* (Female-specific) and *RPS4Y1* / *DDX3Y* (Male-specific) dominate the top log2 fold changes across all cell types, serving as an internal positive control.
  * **Autosomal Signal:** When sex chromosomes are excluded, **CD14+ Monocytes** retain notable expression shifts in inflammatory markers (*S100A8*, *S100A9*, *LYZ*).
* **Donor Consistency:** As shown in Figure 5, both male donors (Donor1, Donor2) show consistent lower expression of *XIST* and similar autosomal baseline trajectories compared to female donors (Donor3, Donor4).
* **Caveat & Confounding Factors:** Because sex (Male vs. Female) is perfectly confounded with age group (18–35 vs. 36–50) across these 4 donors, these autosomal shifts cannot be definitively attributed to sex alone. They represent broader baseline immune variations driven by combined age, sex, and individual donor effects.
